---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```
# bigtime: Sparse Estimation of Large Time Series Models
<!-- badges: start -->
[![CRAN_Version_Badge](http://www.r-pkg.org/badges/version/bigtime)](https://cran.r-project.org/package=bigtime)
[![CRAN_Downloads_Badge](https://cranlogs.r-pkg.org/badges/grand-total/bigtime)](https://cran.r-project.org/package=bigtime)
[![License_GPLv2_Badge](https://img.shields.io/badge/License-GPLv2-yellow.svg)](https://www.gnu.org/licenses/old-licenses/gpl-2.0.html)
[![License_GPLv3_Badge](https://img.shields.io/badge/License-GPLv3-yellow.svg)](https://www.gnu.org/licenses/gpl-3.0.html)
<!-- badges: end -->

The goal of `bigtime` is to sparsely estimate large time series models such as the Vector AutoRegressive (VAR) Model, the Vector AutoRegressive with Exogenous Variables (VARX) Model, and the Vector AutoRegressive Moving Average (VARMA) Model. The univariate cases are also supported.

## Installation

You can install bigtime from github a follows:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("ineswilms/bigtime")
```

## Plotting the data
We will use the time series contained in the `example` data set. The first ten columns in our dataset are used as endogenous time series in the VAR and VARMA models, and the  last five columns are used as exogenous time series in the VARX model. Note that we remove the last observation from our dataset as we will use this one to illustrate how to evaluate prediction performance. We start by making a plot of our data.

```{r}
library(bigtime)
suppressMessages(library(tidyverse)) # Will be used for nicer visualisations

data(example)
Y <- example[-nrow(example), 1:10] # endogenous time series
colnames(Y) <- paste0("Y", 1:ncol(Y)) # Assigning column names
Ytest <- example[nrow(example), 1:10]
X <- example[-nrow(example), 11:15] # exogenous time series
colnames(X) <- paste0("X", 1:ncol(X)) # Assinging column names


plot_series <- function(Y){
  as_tibble(Y) %>%
  mutate(Time = 1:n()) %>%
  pivot_longer(-Time, names_to = "Series", values_to = "vals") %>%
  ggplot() +
  geom_line(aes(Time, vals)) + 
  facet_wrap(facets = vars(Series), ncol = 2) + 
  ylab("") +
  theme_bw()
}

plot_series(Y)
plot_series(X)
```

## Multivariate Time Series Models

### Vector AutoRegressive (VAR) Models

While we could use the example data provided, *bigtime* also support simulation of VAR models using both lasso and elementwise type sparsity patterns. Since lasso is the most widely known one, we will start off with lasso. To simulate a VAR model having lasso type of sparsity, use the `simVAR` function and set `sparsity_pattern="lasso"`. The lasso sparsity patterns the addional `num_zero` options which determines the number of zeros in the VAR coefficient matrix (excluding intercepts). *Note: we also set a seed so that the simulation is replicatable*. We can then use the `summary` function to obtain a summary of the simulated data.

```{r}
periods <- 200
k <- 5 # Number of time series
p <- 5 # Maximum lag 
sparsity_options <- list(num_zero = 15)
sim_data <- simVAR(periods = periods, k = k, p = p, 
                   sparsity_pattern = "lasso", 
                   sparsity_options = sparsity_options,
                   seed = 123456, 
                   decay = 0.01) # the smaller decay the larger early coefs w.r.t. to later ones
Y <- sim_data$Y
summary(sim_data) # Obtaining a summary of the simulation
```

The above simulated data (and truly any other data) could then be estimated using an L1-penalty (lasso penalty) on the autoregressive coefficients. To this end, set the `VARpen` argument in the `sparseVAR` function equal to L1. 

For the selection of the penalisation parameter, we can either set the `selection` argument to `"none"`, which would return a model for a sequence of penalisations, or use time series cross-validation by setting `selection="cv"`, or finally we could also use any of BIC, AIC, or HQ information criteria by setting the `selection` arguments to any of `"bic"`, `"aic"`, or `"hq"` respectively. We will start of by using time series cross-validation and will therefore set `selection="cv"`. The default  is to use a cross-validation score based on one-step ahead predictions but you can change the default forecast horizon under the argument `h`. *Note: it is recommended to standardise the data. Bigtime will give a warning if the data is not standardised but will not stop you from continuing.*

```{r}
VARL1 <- sparseVAR(Y=scale(Y), 
                   selection = "cv", 
                   VARpen = "L1") # Using scale() to standardise data
```

The `plot_cv` function can be used to investigate the cross-validation procedure. The black line indicates the penalty parameter choice that lead to the smallest MSFE in the CV procedure. The red line, on the other hand, shows the one SE optimal penalisation. The latter is the one that is chosen by `sparseVAR`.

```{r}
plot_cv(VARL1)
```

Further investigation into the model can be done by using the function `lagmatrix`, which returns the lagmatrix of the estimated autoregressive coefficients. If entry $(i,j)=x$, this means that the sparse estimator indicates the effect of time series $j$ on time series $i$ to last for $x$ periods. Setting the `returnplot` argument to `TRUE` will return a heatmap for better visual inspection. 

```{r}
LhatL1 <- lagmatrix(fit=VARL1, returnplot=TRUE)
```

The lag matrix is typically sparse as it contains some empty (i.e., zero) cells. However, VAR models estimated with a standard L1-penalty are typically not easily interpretable as they select many high lag order coefficients (i.e., large values in the lagmatrix). 

To circumvent this problem, we advise using a lag-based hierarchically sparse estimation procedure, which boils down to using the default option HLag for the `VARpen` argument.
This estimation procedures encourages low maximum lag orders, often results in sparser lagmatrices, and hence more interpretable models.

A VAR with HLag (elementwise) type of sparsity can be simulated using `simVAR` and by setting `sparsity_pattern="hvar"`. Three extra options exist: (1) `zero_min` determines the minimum number of zero coefficients of each variable in each equation, (2) `zero_max` determines the maximum number of zero coefficients of each variable in each equation, and (3) `zeroes_in_self` is a boolean option that if TRUE, zero coefficients of the $i$th variable can exist in the $i$th equation.

```{r}
periods <- 200
k <- 5 # Number of time series
p <- 5 # Maximum lag 
sparsity_options <- list(zero_min = 0, 
                         zero_max = 5, 
                         zeroes_in_self = TRUE)
sim_data <- simVAR(periods = periods, k = k, p = p, 
                   sparsity_pattern = "hvar", 
                   sparsity_options = sparsity_options,
                   seed = 123456, 
                   decay = 0.01)
Y <- sim_data$Y
summary(sim_data) # Obtaining a summary of the simulation
```
Models can be estimated using the hierarchical penalisation by using the default argument to `VARpen`, namely `HLag`. Model selection can again be done by either setting `selection="none"` and obtaining a whole sequence of models, or by using any of `cv, bic, aic, hq`. 


```{r}
VARHLag_none <- sparseVAR(Y=scale(Y), selection = "none") # HLag is the dafault VARpen option
VARHLag_cv <- sparseVAR(Y=scale(Y), selection = "cv")
VARHLag_bic <- sparseVAR(Y=scale(Y), selection = "bic") # This will also give a table for IC comparison
```
Depending on which selection procedure was used, various diagnostics can be produced. Former and foremost, all selection procedures support the `fitted` and `residuals` functions to obtain the fitted and residual values respective. Both functions return a 3D array if the model used `selection="none"` correpoding to the fitted/residual values for each model in the penalisation sequence.

```{r}
res_VARHLag_none <- residuals(VARHLag_none) # This is a 3D array
dim(res_VARHLag_none)
```

When an actual selection method was used (`cv, bic, aic, hq`), then other diagnostic functions exist. For `cv`, `plot_cv` could be used again, just as shown above. For all, the `diagnostics_plot` function can be used to obtain a plot of fitted and residual values. 

```{r}
p_bic <- diagnostics_plot(VARHLag_bic, variable = "Y3") # variable argument can be numeric or character
p_cv <- diagnostics_plot(VARHLag_cv, variable = "Y3") # variable argument can be numeric or character

plot(p_bic)
plot(p_cv)
```

### Vector AutoRegressive with Exogenous Variables (VARX) Models

```{r}
data(example)
Y <- example[-nrow(example), 1:10] # endogenous time series
colnames(Y) <- paste0("Y", 1:ncol(Y)) # Assigning column names
Ytest <- example[nrow(example), 1:10]
X <- example[-nrow(example), 11:15] # exogenous time series
colnames(X) <- paste0("X", 1:ncol(X)) # Assinging column names
```

Often practitioners are interested in incorparating the impact of unmodeled exogenous variables (X) into the VAR model. To do this, you can use the `sparseVARX` function which has an argument `X` where you can enter the data matrix of exogenous time series.

When applying the `lagmatrix` function to an estimated sparse VARX model, the lag matrices of both the endogenous and exogenous autoregressive coefficients are returned.

`sparseVARX` supports the same `selection` arguments as `sparseVAR`: `none, cv, bic, aic, hq`, and it is again recommended to standardise the data.

```{r}
VARXfit_cv <- sparseVARX(Y=scale(Y), X=scale(X), selection = "cv") # VARX models always to Cross-Validation
LhatVARX <- lagmatrix(fit=VARXfit_cv, returnplot=TRUE)
```

VARX models also support `plot_cv` if estimated using CV. The big black dot in the plot below indicates the SE optimal choice, while the contours indicate the mean MSFE in the CV procedure.

```{r}
plot_cv(VARXfit_cv)
```


If `selection="none"` a 3D array will be returned again. Although not mentioned previously, when setting `selection` to `none`, or any of the IC, one can also easily provide a penalisation sequence, or even just ask for a single penalisation setting. For example, below we intentionally choose a single, small penalisation for the exogenous variables.

```{r}
VARXfit_none <- sparseVARX(Y=scale(Y), X=scale(X), VARXlBseq = 0.001, selection = "none")
dim(VARXfit_none$Phihat) # This is a 3D array
# This is also 3D but third dimension is equal to ten 
# --> one penalisation was chosen for B and 10 automatically for Phi
# --> Cross product makes 10
dim(VARXfit_none$Bhat) 
```

Other functions such as `residuals`, `fitted`, and `diagnosticsplot` are also supported.

### Vector AutoRegressive Moving Average (VARMA) Models

VARMA models generalized VAR models and often allow for more parsimonious representations of the data generating process. To estimate a VARMA model to a multivariate time series data set, use the function `sparseVARMA`, and choose a desired selection method. As a default, `sparseVARMA` uses CV in the first stage and `none` in the second stage. **The first stage does not support `none`: A selection needs to be made.**

Now lag matrices are obtained for the autoregressive (AR) coefficients and the moving average (MAs) coefficients.
```{r}
VARMAfit <- sparseVARMA(Y=scale(Y), VARMAselection = "cv") # VARselection="cv" as default.  
LhatVARMA <- lagmatrix(fit=VARMAfit, returnplot=T)
```

Other functions such as `plot_cv`, `residuals`, `fitted` and `diagnosticsplot` are also supported.

## Evaluating Forecast Performance

To obtain forecasts from the estimated models, you can use the `directforecast` function. The default forecast horizon (argument `h`) is set to one such that one-step ahead forecasts are obtained, but you can specify your desired forecast horizon. Finally, we compare the forecast accuracy of the different models by comparing their forecasts to the actual time series values.
In this example, the VARMA model has the best forecast performance (i.e., lowest mean squared prediction error). This is no surprise as the multivariate time series  $Y$ was generated from a VARMA model.
```{r}
# VARf <- directforecast(VARHLag) # default is h=1
# VARXf <- directforecast(VARXfit)
# VARMAf <- directforecast(VARMAfit)
# 
# mean((VARf-Ytest)^2)
# mean((VARXf-Ytest)^2)
# mean((VARMAf-Ytest)^2) # lowest=best
```

## Univariate Models
The functions `sparseVAR`, `sparseVARX`, `sparseVARMA` can also be used for the univariate setting where the response time series $Y$ is univariate. Below we illustrate the usefulness of the sparse estimation procedure as automatic lag selection procedures.


### AutoRegressive (AR) Models
We start by generating a time series of length $n=50$ from a stationary AR model and by plotting it. The `sparseVAR` function can also be used in the univariate case as it allows the argument `Y` to be a vector. The `lagmatrix` function gives the selected autoregressive order of the sparse AR model. The true order is one.
```{r}
# n <- 50
# y <- rep(NA, n)
# y[1] <- rnorm(1, sd=0.1)
# for(i in 2:n){
#   y[i] <- 0.5*y[i-1] + rnorm(1, sd=0.1)
# }
# par(mfrow=c(1,1))
# plot(y, type="l", xlab="Time", ylab="")
# ARfit <- sparseVAR(Y=y, selection = "cv")
# lagmatrix(fit=ARfit)
```

### AutoRegressive with Exogenous Variables (ARX) Models
We start by generating a time series of length $n=50$ from a stationary ARX model and by plotting it. The `sparseVARX` function can also be used in the univariate case as it allows the arguments `Y` and `X` to be vectors. The `lagmatrix` function gives the selected endogenous (under `LPhi`) and exogenous autoregressive (under `LB`) orders of the sparse ARX model. The true orders are one.
```{r}
# n <- 50
# y  <- x <- rep(NA, n)
# x[1] <- rnorm(1, sd=0.1)
# y[1]  <- rnorm(1, sd=0.1)
# for(i in 2:n){
#   x[i] <- 0.8*x[i-1] + rnorm(1, sd=0.1)
#   y[i] <- 0.5*y[i-1] + 0.2*x[i-1] + rnorm(1, sd=0.1)
# }
# par(mfrow=c(1,1))
# plot(y, type="l", xlab="Time", ylab="")
# ARXfit <- sparseVARX(Y=y, X=x, selection = "cv") 
# lagmatrix(fit=ARXfit)
```


### AutoRegressive Moving Average (ARMA) Models
We start by generating a time series of length $n=50$ from a stationary ARMA model and by plotting it. The `sparseVARMA` function can also be used in the univariate case as it allows the argument `Y` to be a vector. The `lagmatrix` function gives the selected autoregressive (under `LPhi`) and moving average  (under `LTheta`) orders of the sparse ARMA model. The true orders are one.
```{r}
# n <- 50
# y <- u <- rep(NA, n)
# y[1] <- rnorm(1, sd=0.1)
# u[1] <- rnorm(1, sd=0.1)
# for(i in 2:n){
#   u[i] <- rnorm(1, sd=0.1)
#   y[i] <- 0.5*y[i-1] + 0.2*u[i-1] + u[i]
# }
# par(mfrow=c(1,1))
# plot(y, type="l", xlab="Time", ylab="")
# ARMAfit <- sparseVARMA(Y=y, VARMAselection = "cv") 
# lagmatrix(fit=ARMAfit)
```

## Additional Resources

For a non-technical introduction to VAR models see this interactive [notebook](https://mybinder.org/v2/gh/enweg/SnT_VARS/main?urlpath=shiny/App/) and for an interactive notebook demonstrating the use of BigTime for high-dimensional VARs, see this  [notebook](https://mybinder.org/v2/gh/enweg/SnT_BigTime/main?urlpath=shiny/App/). *Note: Loading these notebooks sometimes can take quite some time. Please be patient or try another time.*


## References:

* Nicholson William B., Bien Jacob and Matteson David S. (2020), "High-dimensional forecasting via interpretable vector autoregression", Journal of Machine Learning Research, 21(166), 1-52.

* Wilms Ines, Sumanta Basu, Bien Jacob and Matteson David S. (2017), "Sparse Identification and Estimation of High-Dimensional Vector AutoRegressive Moving Averages", arXiv:1707.09208.

* Wilms Ines, Sumanta Basu, Bien Jacob and Matteson David S. (2017), "Interpretable Vector AutoRegressions with Exogenous Time Series", arXiv.
